# Describing continuous variables

How do you summarize a continuous variable and present your findings? Here we cover how to analyze data over a single continuous variable, both in isolation (e.g. basic summary statistics) and across discrete categories of another variable (e.g. group mean comparison).

You will need the `dcps` data (`DCPS testing.RData`) and the `tidyverse` package to replicate the examples in this chapter. See Section \@ref(import) for a reminder on how to import the data and Section \@ref(packages) for installing and loading packages.
``` {r chapterload, echo=F, message=F}
library(tidyverse)
load('docs/data/DCPS testing.RData')
```

## Describing one variable

### Summary statistics {#contsumstat}
A useful first step in analyzing the distribution of scores on a single numeric variable is to calculate the relevant summary statistics. Use the `summary()` function for a quick, general overview. This returns the minimum, mean, and maximum scores, as well as the score at 1st, 2nd (median), and 3rd quartiles.

For specific inquiries, use the `summarize()` function and customize your report. For example:

```{r summarise, eval=TRUE}
  library(tidyverse) # load tidyverse package for pipe syntax and summary functions
  dcps %>%   # start by piping in the dataset
    summarize(
      Avg = mean(ProfLang), # calculates the mean
      StdDev = sd(ProfLang),   # standard deviation
      Range = max(ProfLang) - min(ProfLang) 
    )
```

### Graphing the distribution
In addition to collapsing the data to summary statistics, it can be informative to plot the distribution of a variable, i.e. displaying the probability of observing different values or ranges of values. Chapter \@ref(ggplot) introduces plotting generally and Section \@ref(ggplothist) covers techniques for plotting distributions of a single variable.

## Omitting missing values

If some observations have missing data (coded `NA`) on a given variable, summary statistic functions may yield an error. To ignore any missing values in carrying out calculations, include the argument `na.rm=TRUE` in the function.

```{r nas, eval=TRUE}
  a = c(1,3,5,NA,7)

  mean(a) # Operation fails
  
  mean(a, na.rm = TRUE) # Successfully calculates the mean
```


## Group comparisons

### Summary statistics by group {#contsumstatgroup}
Comparing outcomes across groups can reveal important patterns. We conduct such comparisons using the `group_by()` function, here in combination with `summarize()`. The `group_by()` function takes as arguments a comma separated list variables. R will then evaluate all piped commands separately for every combination of values that appear in your listed variables. Here's an example that calculates the mean and standard deviation of math proficiency (`ProfMath`) separately for elementary, middle, and high school (`SchType`).

```{r grpby, eval=TRUE, warning=FALSE, message=FALSE}
  dcps %>%
    group_by(SchType) %>%  # separately for each value of SchType
    summarize(
      Avg = mean(ProfMath),  # calculate mean of ProfMath
      StDev = sd(ProfMath)   # calculate SD of ProfMath
    )
```

Note: the arguments for `group_by()` should be categorical variables and not continuous variables. Why? What happens when you try using a continuous variable?

### Visualize group differences
Plotting distributions separately for different groups can offer insights into the relationships between variables. Chapter \@ref(ggplot) introduces plotting generally and Section \@ref(ggplotrelate) covers techniques for visualizing group differences and relationships between variables.
