# Describing categorical variables

Categorical variables require a different approach. Rather than summary statistics, the relevant information here is the frequency with which each value or category appears in the data. Here we cover how to analyze data for a single categorical variable (e.g. frequency tables and bar charts) and how to evaluate relationships between nominal variables (cross-tabulation).

Note that you will need the `film` data (`biopics.xls`) and the `tidyverse` package to replicate the examples in this chapter. See Section \@ref(import) for a reminder on how to import the data and Section \@ref(packages) for installing and loading packages.
``` {r chapterload, echo=F, message=F}
library(tidyverse)
library(readxl)
film <- read_excel('biopics.xls')
```

## Describing one variable {#tab}

To calculate the frequency of different values of a categorical variable, we adapt the combination of `group_by()` and `summarize()` from Section \@ref(contsumstatgroup). In this example, we count the number of biopics that featured women vs. men, as captured by the variable `SubjectSex`. First, we use the command `group_by(SubjectSex)` to instruct R that all subsequent commands should be evaluated separately for different values of `SubjectSex`. Then we use `summarize()` to create a custom set of summary statistics. Finally, we use the function `n()` to count observations, and we set the result equal to a new variable `Count`.

```{r rawtab, eval=TRUE, message=FALSE}

# Frequency table
  Tab <- film %>% # create new object 'Tab' starting with object 'film'
    group_by(SubjectSex) %>% # separately for each value of SubjectSex
    summarize(Count=n()) # count the number of observations
  
  Tab # print object 'Tab' to the console
```

Calculating the the fraction of total cases in each category (relative frequency) requires an extra line of code (`mutate(Percent = Count/sum(Count))`). The `mutate()` function creates a new variable (see Section \@ref(mutate) for details). The argument starts with the name of the new variable (`Percent`) and a definition of the new variable. In this case, we set `Percent` equal to the value of `Count` in each row divided by the sum of the values of `Count` across all rows. 

```{r reltab, eval=TRUE}
# Relative frequency
  Tab_pct <- Tab %>% # create new object 'Tab_pct' starting with object 'Tab'
    mutate(Percent = Count/sum(Count)) # create new variable 'Percent'
      # and set it equal to the value of Count in each row divided by
      # the value of Count summed across all rows, i.e. relative frequency
  
  Tab_pct # print object 'Tab_pct' to the console
```

Frequency tables can also be visualized as bar charts. See Chapter \@ref(ggplot) for plotting generally and Section \@ref(ggplothist) for examples of a bar chart. Frequency tables and bar charts are two different ways of representing the distribution of a categorical variable.

## Cross-tabulation
We can examine the relationship between two categorical variables with a simple extension of the code from Section \@ref(tab). In the example below, we see if the gender of the subject varies (`SubjectSex`) by the period of the film (`Period`). To do so, we simply add the variable `Period` to the `group_by()` function that we used in the previous Section:

```{r xtabct, eval=TRUE}

# Frequency table
  XTab <- film %>% # create new object 'XTab' starting with object 'film'
    group_by(SubjectSex,Period) %>% # separately for every combination of values
                                    # of 'SubjectSex' and 'Period'
    summarize(Count=n()) # count the number of observations
  
  XTab # print object 'XTab' to the console
```

In this table, the comparison we want to make, whether the gender focus of biopics has changed over time, is somewhat obscured by the change in the total number of biopics over time. We can again adapt code from Section \@ref(tab) to calculate the percentage of films with a subject of each gender separately for each period.

```{r xtabpct, eval=TRUE}
# Frequency table with percentages by Period
  XTabPct <- XTab %>% # create new object 'XTabPct' starting with object 'XTab'
    group_by(Period) %>% # separately for every value of 'Period'
    mutate(PercentPeriod = Count/sum(Count)) # create new variable
      # 'PercentPeriod' and set it equal to the value of Count in each row 
      # divided by the value of Count summed across all rows (within a value of 
      # 'Period'), i.e. relative frequency by period.

  XTabPct # print object 'XTabPct' to the console
```

We see that the fraction of biopics with female subjects is relatively constant over time, between 22.5 and 25 percent. The function `tabyl()` in the `janitor` package is a simple but powerful alternative for producing customized frequency tables.

## Means of binary variables
In most cases, calculating summary statistics for categorical variables does not yield interpretable results. As an example, imagine we asked 5 Dutch people how they commuted on this morning. Their answers include biking, walking, and the bus. The result is a categorical variable that we could store in several different ways.

- As text, with the strings `Bike`,`Walk`, and `Bus` indicating the different categories
- As numbers, with `0` corresponding to biking, `1` to walking, and `2` to the bus
- As numbers, with `1` corresponding to biking, `2` to walking, and `3` to the bus

We could implement each of these coding schemes in R, creating a new object with 5 observations and 3 variables:

```{r categorical, eval=TRUE}
commute <- tibble( # create a new object 'commute' defined as a tidyverse tibble
                  text = c("Bike","Bike","Walk","Walk","Bus"), # variable 'text' stores commute as text
                  code1 = c(0,0,1,1,2), # variable 'code1' stores commute as numerical codes 0,1,2
                  code2 = c(1,1,2,2,3)  # variable 'code2' stores commute as numerical codes 1,2,3
                  )

commute # print object 'commute' to console
```

All 3 of these variables contain *exactly the same information*, just stored differently. What if we tried to take the means of each?


```{r mean_text, eval=TRUE, warning = FALSE, message = FALSE}
commute %>%
  summarize(mean_text = mean(text), # create a new variable 'mean_text' equal to the mean of the variable 'text'
            mean_code1 = mean(code1), # create a new variable 'mean_code1' equal to the mean of the variable 'code1'
            mean_code2 = mean(code2)) # create a new variable 'mean_code2' equal to the mean of the variable 'code2'
```

The results are not terribly useful. Arithmetic operations are not defined for text, so we get a missing value for the mean of `text`. Since the other variables use a numerical coding scheme, it is possible to calculate a mean, but the results of 0.8 and 1.8 have two obvious deficiencies. First, they are different from each other even though the underlying information is exactly the same. Second, they are not comprehensible answers to the question "what modes of commuting are typical for our Dutch survey respondents?"

But let's consider one more way of representing the data: a series of binary categorical variables, one for each mode of commuting. These variables will share a coding scheme, assuming a value of 0 if the commuter did not use that mode and a value of 1 if the commuter did use that mode. Here's what the data would look like under that coding scheme.

```{r cat_binary, eval=TRUE}
commute_binary <- tibble( # create a new object 'commute_binary' defined as a tidyverse tibble
                  bike = c(1,1,0,0,0), # variable 'bike' assumes a value of 1 for bike commutes, 0 otherwise
                  walk = c(0,0,1,1,0), # variable 'walk' assumes a value of 1 for walking commutes, 0 otherwise
                  bus = c(0,0,0,0,1)  # variable 'bus' assumes a value of 1 for bus commutes, 0 otherwise
                  )

commute_binary # print object 'commute_binary' to console
```

This table again stores exactly the same information, but now we can learn something from the means.

```{r mean_binary, eval=TRUE, warning = FALSE, message = FALSE}
commute_binary %>%
  summarize(mean_bike = mean(bike), # create a new variable 'mean_bike' equal to the mean of the variable 'bike'
            mean_walk = mean(walk), # create a new variable 'mean_walk' equal to the mean of the variable 'walk'
            mean_bus = mean(bus)) # create a new variable 'mean_bus' equal to the mean of the variable 'bus'
```

These means correspond to the fraction of survey respondents who used each mode of transport; 40 percent biked, 40 percent walked, and 20 percent rode the bus. In general, the mean of an binary variable coded as `(0,1)` will give you the fraction of observations with a value of 1.