# Doing more with data frames

Very few data sets come "ready made" for analysis. Even in our introductory course, there are times when you will need to limit your analysis to certain observations, recode a variable, and more. This chapter covers a few basics of data "munging."

Note that you will need the `film` (`"biopics.xls"`) and `dcps` (`"DCPS testing.RData"`) data sets and `tidyverse` package to replicate the commands.  See Section \@ref(import) for a reminder on how to import the data and Section \@ref(packages) for installing and loading packages.
``` {r chapterload, echo=F, message=F}
library(tidyverse)
library(readxl)
film <- read_excel('docs/data/biopics.xls')
load('docs/data/DCPS testing.RData')
```

## Filter/subset data
It is often necessary to limit your analysis to some subset of observations. Use the `filter()` command to specify the criteria by which to select observations.

```{r filter,eval=TRUE}
  film_fem <- film %>% # create new object 'film_fem' starting with object 'film'
    filter(SubjectSex == 'Female') # Choose observations where the variable
      # 'SubjectSex' has a value of 'Female'

  film_fem # Print 'film_fem' to the console
```

The conditions inside `filter()` identify the observations, or rows, to keep (i.e. you're selecting only those rows that satisfy the given conditions). This can be based on any number of conditions. Note that a double equals sign `==` is used to *check* equality while a single equals sign `=` is used to *assign values*. Since a `filter()` function is designed to evaluate criteria for each observation, rather than assign values, we will always use a double equals sign `==` with `filter()`.

Filtering or subsetting by a text-based variable like `SubjectSex` requires that you enclose the desired value in single or double quotation marks, i.e. `SubjectSex == 'Female'` or `SubjectSex == "Female"`, not `SubjectSex == Female`. Without quotation marks, R will look for an *object* named `Female`. However, if you want to filter a numeric variable, you will generally omit the quotation marks. For instance, say we are interested in films releasted in 2014:

```{r filter_year,eval=TRUE}
  film_14 <- film %>% # create new object 'film_14' starting with object 'film'
    filter(Release == 2014) # Choose observations where the variable
      # 'Release' has a value of 2014

  film_14 # Print 'film_14' to the console
```


## Selecting your variables
A dataframe with loads of variables can be unwieldy. Use the `select()` function to isolate the variables you need for your analysis.

```{r select,eval=TRUE}
  Study1 <- film %>%
    select(Title,Release,SubjectSex)

  head(Study1)
```

## Create new variables {#mutate}
The `mutate()` function creates new variables (or columns) in a data frame. By assigning the same object name to the result, the new variables become part of the object.

```{r mutate,eval=TRUE}
  dcps <- dcps %>%
    mutate(
      LastEdit = '2021',  # character variable with same value in each
      ProfLangProp = ProfLang/100, # convert to proportion
    )
  
  dcps %>% select(2,5,8:9) %>% head()
```

In addition to transformations that apply the same function to every case, you can create variables that treat observations on an individual basis. Consider the `if_else()` option. This will specify a logical test to apply to each case, and specify replacement values for those observations that evaluate as true and those observations that evaluate as false. The syntax is: `if_else(criteria,value.if.true,value.if.false)`.

```{r ifelse,eval=TRUE}
  dcps <- dcps %>%
    mutate(
      AbvMedMath = if_else(
        ProfMath > median(ProfMath), # condition to evaluate
        1,  # output if condition is TRUE
        0   # output if FALSE
      )
    )

  dcps %>% select(SchCode,ProfMath,AbvMedMath) %>% head()
```

## Appending and merging data

It is often useful to combine data from different sources. This may take the form of appending (adding additional observations with information on the same variables) or merging (adding additional variables that describe the same observations).

### Appending new observations
To append new observations to your data frame, use `bind_rows(OldData,NewData)`. Note that the variable names need to match exactly across data frames, but the variable order does not matter.

```{r append,eval=TRUE}
  # old data
    myData = tribble(
      ~District, ~Students,
      115, 985,
      116, 1132
    )

  # new data to add
    new = tribble(
      ~District, ~Students,
      117, 419,
      118, 633
    )
    
  # Append new to old
    myData = bind_rows(myData,new)
    
    myData
```

### Merging
To merge data frames (add new variables for existing observations), use `left_join(OldData,NewData)`. In order to link rows in one data frame to rows in another, it is critical that the data sets contain a common identifier, with the same variable name and same values. Building on the example above

```{r merge,eval=TRUE}
  # new variables
    newvars = tribble(
      ~District,~Teachers,
      115, 43,
      116, 71,
      118, 55
    )

  # join new to old
    myData = left_join(myData,newvars)
    
    myData
```
